---
title: "Dead cell analysis - all samples"
author: "Elisa Roesti"
date: '2022-07-28'
output: md_document
---

```{r, warning=FALSE, message=FALSE}
library(Seurat)
library(dplyr)
library(SingleCellExperiment)
library(cowplot)
library(sccomp)
library(ggplot2)
library(patchwork)
library(introdataviz)
library(clusterProfiler)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(dittoSeq)
library(tidyseurat)
library(tidyr)
library(tidyverse)
library(tidySingleCellExperiment)
library(tidysc)
```


## Loading all samples & define alive/dead


```{r}
#select directory
files <- dir("/stornext/Bioinf/data/bioinf-data/Papenfuss_lab/projects/roestie/roestielisa/PBMC/data/3_prime_batch_1/preprocessing_results/non_batch_variation_removal/", full.name = T)
#combine count dataframe by rows using seurat function merge
merged_combined <- 
  files |>
  map(readRDS) |>
  purrr::reduce(merge)

merged_combined #98,626 × 16
```

```{r}
#loading, merging and left_join
files_annotation <- dir("/stornext/Bioinf/data/bioinf-data/Papenfuss_lab/projects/roestie/roestielisa/PBMC/data/3_prime_batch_1/preprocessing_results/annotation_label_transfer/", full.name = T)
files_alive_identification <- dir("/stornext/Bioinf/data/bioinf-data/Papenfuss_lab/projects/roestie/roestielisa/PBMC/data/3_prime_batch_1/preprocessing_results/alive_identification", full.name = T)

```



```{r}
#load merge and join:
merged_combined_annotation_alive <- #after annotation 98,626 × 20, after alive: 98,626 × 29
  
  #join annotation
  merged_combined |>
  left_join(
    files_annotation |>
      map(readRDS) |>
      purrr::reduce(bind_rows), by=".cell") |>
  #join alive identification
  left_join(
    files_alive_identification |>
      map(readRDS) |>
      purrr::reduce(bind_rows), by = c(".cell", "predicted.celltype.l1", "refUMAP_1", "refUMAP_2"))

```

```{r}
merged_combined_annotation_alive %>% select(sample) %>% table()
```



```{r, warning=FALSE, message=FALSE, echo=FALSE}
merged_combined_annotation_alive %>% select(alive) %>% table() #FALSE6441  TRUE 92185
#select RNA data
DefaultAssay(merged_combined_annotation_alive) <- "SCT"
```


### Identity ?
```{r, warning=FALSE, message=FALSE, echo=FALSE}
Idents(object = merged_combined_annotation_alive) <- "sample"
#Idents(object = merged_combined_annotation_alive)
```



## Plot 1 - Visualisation of Mitochondria content

### For each sample

```{r}
#original ggplot
merged_combined_annotation_alive |>
  
  #violin plot
  ggplot(aes(x=sample, y=subsets_Mito_percent, fill=alive)) + 
  introdataviz::geom_split_violin() +
  scale_y_sqrt() +
  theme(axis.text.x=element_text(angle=70, hjust=1)) +
  theme_bw()
```


### For each sample according to cell type

```{r, fig.height= 50, fig.width= 20}
#facet by sample, x= celltypes
merged_combined_annotation_alive |>
  
  #violin plot
  ggplot(aes(x=predicted.celltype.l1, y=subsets_Mito_percent, fill=alive)) + 
  introdataviz::geom_split_violin() +
  facet_wrap(~sample, scale="free_y", ncol =1) +
  theme(axis.text.x=element_text(angle=70, hjust=1)) +
  theme_bw()
```

## Plot 2 - Visualisation of apoptosis, necrosis, autophagy and stress markers

```{r, warning=FALSE, message=FALSE}
# selecting stress markers - splitting the markers for better visualisation
stress_markers <- c("JUN", "FOS", "IL6", "TNF", "CXCR4", "SNAI1", "VIM", "GADD45B", "MCL1","STAT1", "IRF7", "IRF3", "IRF1", "CDKN2A", "TP53", "MRTFB", "RBL1", "NUPR1", "IFNG", "IFNB1", "IFNA2", "IFNL1", "IFNA1", "IFNK", "TNF", "IL-12")

#selecting death markers
apoptosis_markers <- c("BCL2L1", "CASP9", "CHP2", "CYCS", "EXOG", "IL1A", "IL1R1", "IL1RAP", "IL3RA", "PIK3CA", "PIK3CD", "PIK3CG", "PIK3R1", "PIK3R2", "PRKAR1B", "NFKBIA", "TNFRSF10A", "TNFRSF10B", "TNFRSF10D", "TNFRSF1A")
necrosis_markers = c("DNML1", "GSDME", "IPMK", "MLKL", "RBCK1", "TICAM1", "YBX3")
autophagy_markers = c("ATG12", "GABARAPL1", "IFNA17", "IFNA8")
```



```{r}
#first drop the MT gene from the nCount_SCT :
mitochondrial_genes = grep("^MT-", rownames(merged_combined_annotation_alive[["SCT"]]),value = T)
ribosomal_genes = grep("^RP(S|L)", rownames(merged_combined_annotation_alive[["SCT"]]), value = T)

#remove the gene from the matrix #dim before 98,626 × 29.     98,626 × 29
merged_combined_annotation_alive_no_MT_no_rib <-
  merged_combined_annotation_alive[! rownames(merged_combined_annotation_alive) %in% mitochondrial_genes, ]

merged_combined_annotation_alive_no_MT_no_rib <-
  merged_combined_annotation_alive_no_MT_no_rib[! rownames(merged_combined_annotation_alive_no_MT_no_rib) %in% ribosomal_genes, ]
```


```{r}
#re-scaling, see github
all.genes <- rownames(merged_combined_annotation_alive_no_MT_no_rib)
head(all.genes)
merged_combined_annotation_alive_no_MT_no_rib<- ScaleData(merged_combined_annotation_alive_no_MT_no_rib, assay = "SCT", slot='scale.data', features = all.genes)
```


```{r, fig.height=20}
# Creating gene signatures:

# 1 create a gene signature + genes in the signature dataframe:
#stress
stress_markers_df <- as.data.frame(stress_markers)
stress_markers_df$signature <- "stress"
colnames(stress_markers_df)[1] <- ".feature"

apoptosis_markers_df <- as.data.frame(apoptosis_markers)
apoptosis_markers_df$signature <- "apoptosis"
colnames(apoptosis_markers_df)[1] <- ".feature"

necrosis_markers_df <- as.data.frame(necrosis_markers)
necrosis_markers_df$signature <- "necrosis"
colnames(necrosis_markers_df)[1] <- ".feature"

autophagy_markers_df <- as.data.frame(autophagy_markers)
autophagy_markers_df$signature <- "autophagy"
colnames(autophagy_markers_df)[1] <- ".feature"

#bind by rows, obtain only 1 dataframe with all the signatures
gene_signatures <- rbind(stress_markers_df, apoptosis_markers_df, necrosis_markers_df, autophagy_markers_df)


#combine the new dataframe with my seurat object:
#join seurat object with signature dataframe, option long, bind by col created = .features
merged_combined_annotation_alive_no_MT_no_rib_signature <-
merged_combined_annotation_alive_no_MT_no_rib |>
  join_features(
    features = c(stress_markers, apoptosis_markers, necrosis_markers, autophagy_markers),
    shape="long") |>
  left_join(gene_signatures, by = ".feature")

#Plot 1 - facet by gene signatures
merged_combined_annotation_alive_no_MT_no_rib_signature |>
  group_by(.cell, signature, alive, sample) |>
  mutate(sum_of_counts = sum(.abundance_SCT)) |>
#create the ggplot facet by gene signatures
  ggplot(aes(x=sample, y=sum_of_counts, fill=alive)) + 
  introdataviz::geom_split_violin() +
  facet_wrap(~signature,  scale="free_y" ) + #ncol=1
  theme(axis.text.x=element_text(angle=70, hjust=1)) +
  theme_bw()

#Plot 2 - facet by samples
#create the column sum of counts
merged_combined_annotation_alive_no_MT_no_rib_signature |>
  group_by(.cell, signature, alive, sample) |>
  mutate(sum_of_counts = sum(.abundance_SCT)) |>

#create the ggplot facet by samples
  ggplot(aes(x=signature, y=sum_of_counts, fill=alive)) + 
  introdataviz::geom_split_violin() +
  facet_wrap(~sample, scale="free_y", ncol=3) + #ncol=1
  theme(axis.text.x=element_text(angle=70, hjust=1)) +
  theme_bw()


```




## Plot 3 - Composition of cells

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#merged_samples

# celltype :  predicted.celltype.l1
# counts : nCount_SCT

#I also have to distinguished between alive and dead: I have a label in the column alive_dead, or I can create the sample_alive/dead like below

plot_composition_1 <-
  merged_combined_annotation_alive |>
  unite("sample_alive", c(sample, alive), remove=FALSE) |>
  unite("sample_alive", c(sample, alive), remove=FALSE) |>
  select(sample, nCount_SCT, predicted.celltype.l1, alive, sample_alive) |>
  
  # create frquency column
  mutate(frequency = nCount_SCT/sum(nCount_SCT)*100) |>
  
  # create the proportion column
  group_by(sample_alive) |>
  mutate(tot_sample_proportion = sum(frequency)) |>
  ungroup() |>
  mutate(proportion = (frequency * 1)/tot_sample_proportion) |>
  
  #plot proportion
  ggplot(aes(x = sample_alive, y = proportion, fill = predicted.celltype.l1)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=90, hjust=1)) +
  facet_wrap(~alive)

plot_composition_1
```



```{r, fig.width = 15, fig.height = 8, warning=FALSE, message=FALSE, echo=FALSE}
plot_composition_2 <-
  merged_combined_annotation_alive |>
  unite("sample_alive", c(sample, alive), remove=FALSE) |>
  select(sample, nCount_SCT, predicted.celltype.l2, alive, sample_alive) |>
  
  # create frquency column
  mutate(frequency = nCount_SCT/sum(nCount_SCT)*100) |>
  
  # create the proportion column
  group_by(sample_alive) |>
  mutate(tot_sample_proportion = sum(frequency)) |>
  ungroup() |>
  mutate(proportion = (frequency * 1)/tot_sample_proportion) |>
  
  #plot proportion
  ggplot(aes(x = sample_alive, y = proportion, fill = predicted.celltype.l2)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  facet_wrap(~alive)
  
plot_composition_2
```




```{r, warning=FALSE, message=FALSE, echo=FALSE}
res =
  merged_combined_annotation_alive |>
  
  unite("sample_alive", c(sample, alive), remove=FALSE) |>
  sccomp_glm( 
    formula_composition = ~ alive + sample,  #paired analysis(when adding + sample)
    formula_variability = ~ 1, 
    sample_alive, 
    predicted.celltype.l1
  )
#duplicate rows error when using seurat function to merge samples
```



```{r, warning=FALSE, message=FALSE, echo=FALSE}
res
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#composition plot
plots = plot_summary(res)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#to visualise each cell type in a boxplot
plots$boxplot[1]
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Plot of estimates of differential composition (c_) on the x axis, and differential variability (v_) on the y axis. The error bars represent 95% credible intervals. The dashed lines represent the minimal effect that the hypothesis test is based on. An effect is labelled as significant if bigger than the minimal effect according to the 95% credible interval. Facets represent the covariates in the model.
#plots$credible_intervals_1D
```



## Plot 4 - Check marker for activation induced cell death (Fas/CD95 - FasL/CD95L)

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DefaultAssay(merged_combined_annotation_alive) <- "ADT"
#rownames(merged_samples@assays$ADT@meta.features) # "CD95-Fas" "CD178-Fas-L"
Fas_FasL <- c("CD95-Fas", "CD178-Fas-L")
Fas <- c("CD95-Fas")
FasL <- c("CD178-Fas-L")
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=20}
Fas_plot <-
  merged_combined_annotation_alive |>
  join_features(
    features = Fas,
    shape="long") |> #created a new column .feature , assay="SCT")
  
  #violin plot
  ggplot(aes(x=predicted.celltype.l2, y=.abundance_ADT, fill=alive)) + 
  introdataviz::geom_split_violin(show.legend = FALSE) +
  facet_wrap(~sample, scale="free") +
  theme(axis.text.x=element_text(angle=60, hjust=1)) +
  ggtitle("Fas")

Fas_plot
```


```{r, warning=FALSE, message=FALSE, echo=FALSE,fig.height=20}
Fas_ligand_plot <-
  merged_combined_annotation_alive |>
  join_features(
    features = FasL,
    shape="long") |> #created a new column .feature , assay="SCT") 
  
  #violin plot
  ggplot(aes(x=predicted.celltype.l2, y=.abundance_ADT, fill=alive)) + 
  introdataviz::geom_split_violin() +
  facet_wrap(~sample, scale="free") +
  theme(axis.text.x=element_text(angle=60, hjust=1)) +
  ggtitle("Fas Ligand")

Fas_ligand_plot
```


## Plot 5 - Gene Ontology genes analysis

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(clusterProfiler)
library(AnnotationDbi)
library(org.Hs.eg.db)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Select alive and dead cells as Identities
DefaultAssay(merged_combined_annotation_alive_no_MT_no_rib) <- "RNA"
Idents(merged_combined_annotation_alive_no_MT_no_rib) <- merged_combined_annotation_alive_no_MT_no_rib$alive
levels(Idents(merged_combined_annotation_alive_no_MT_no_rib))
```

### Gene Ontology: Alive cells

```{r, warning=FALSE, message=FALSE, echo=FALSE}
alive.markers <-
  merged_combined_annotation_alive_no_MT_no_rib |>
  NormalizeData(assay="RNA") |>
  ScaleData(assay="RNA") |>
  FindMarkers(
  ident.1 = "TRUE",
  assay = "RNA"
)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE}
#take the positive log2Foldchange genes
alive.markers[alive.markers$avg_log2FC > 0.5, ]
#select gene names to test
gene_to_test_alive <- rownames(alive.markers[alive.markers$avg_log2FC > 0.5, ])
#perform enrichGO
GO_results_BP_alive <- enrichGO(gene = gene_to_test_alive, OrgDb = "org.Hs.eg.db", keyType = "SYMBOL", ont = "BP", pvalueCutoff=0.01) # keytype= "ENTREZID", "ENSEMBL" or "SYMBOL"
```


```{r, fig.width = 5, fig.height = 13, warning=FALSE, message=FALSE, echo=FALSE}
#plot
#fit_alive <- plot(barplot(GO_results_BP_alive, showCategory = 20))
```

### Gene Ontology: Dead cells


```{r, warning=FALSE, message=FALSE, echo=FALSE}
dead.markers <- 
  merged_combined_annotation_alive_no_MT_no_rib |>
  NormalizeData(assay="RNA") |>
  ScaleData(assay="RNA") |>
  FindMarkers(
  ident.1 = "FALSE",
  assay = "RNA",
  min.pct = 0.25
)


```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#take the positive log2Foldchange genes
dead.markers[dead.markers$avg_log2FC > 0.5, ]
#select gene names to test
gene_to_test_dead <- rownames(dead.markers[dead.markers$avg_log2FC > 0.5, ])
#perform enrichGO
GO_results_BP_dead <- enrichGO(gene = gene_to_test_dead, OrgDb = "org.Hs.eg.db", keyType = "SYMBOL", ont = "BP") # keytype= "ENTREZID", "ENSEMBL" or "SYMBOL"
```



```{r, fig.width = 5, fig.height = 13, warning=FALSE, message=FALSE, echo=FALSE}
#plot
fit_dead <- plot(barplot(GO_results_BP_dead, showCategory = 20))
```


